Density-dependent selection in evolutionary genetics: a lottery model of Grime's triangle

### 2017-04-26 manuscript.tex
latest

We assume that the $n_i$ and $T$ are large enough that stochastic fluctuations in the $n_i$ (``drift'') can be ignored. We derive deterministic equations for the expected change in the $n_i$ over time, leaving the evaluation of drift for future work. In particular, we do not evaluate the initial stochastic behaviour of adaptive mutant lineages while they are at low abundances. We restrict our attention to the earliest deterministic behavior of mutant lineages while they are still at low frequency (the transition to deterministic growth occurs at an abundance of order equal to their inverse expected absolute growth rate; \citealt{uecker_2011}).

@@

Stressful environments are more ambiguous, and have been the subject of an extensive debate in the plant ecology literature (the ``Grime-Tilman'' debate; \citealt{aerts_1999} and references therein). Severe stress inhibits growth and reproduction, so that $b\ll 1$ \citep{grime_1974,grime_1977}. In Grime's view, this means that the rate at which propagules successfully develop to adulthood cannot appreciably exceed the mortality rate $b/d\approx 1$. In our model, this implies that territorial contests are rare.

The alternative view is that stressed environments are highly competitive. In particular, if consumable resources are scarce, we expect intense resource competition (for empirical support, see \citealt{davis_1998}). Stress does restrict growth and reproduction, but it also means that fewer individuals can be supported per unit area. Thus, in this view, stressed populations are actually at high density relative the environmental carrying capacity (but low density relative to ideal environments; \citealt{taylor_1990}). Territorial contests are important.

The mapping of different environments to our model parameters is summarized in the first two rows of Fig.~\ref{fig:table}. Also shown is the approximate dependence of $\Delta_+ n_i$ on $b_i$ and $c_i$ for each environment (fourth row). These can be used infer the expected direction of evolution for the traits $b$, $c$ and $d$ (fifth row) using a standard invasion analysis.

We simply invoke the well known result this probability is proportional to $r$, with a proportionality factor typically of order one (Haldane's formula; \citealt{uecker_2011}). The fixation of neutral mutations is exceedingly unlikely (probability of order $1/N$). Consequently, the direction of evolutionary change is determined by the mutational trait changes which are available and also confer an appreciable fitness benefit, where availability is subject to constraints imposed by the environment.

@@

Stable coexistence is possible between genotypes that are superior in different traits. Suppose that $j$ is better at securing territories ($c_j>c_i$), that $i$ is better at producing propagules ($b_i>b_j$), and that $d_i=d_j$. Coexistence occurs if $j$ will invade an $i$-dominated population, but $i$ will also invade a $j$-dominated population (``mutual invasion''). If $b_i$ is so large that $L\gg 1$ when $i$ is dominant, and $b_j$ is so small that $L\ll 1$ when $j$ is dominant, then, combining Eqs. \eqref{eq:equil} and \eqref{eq:invad}, we find that $i$ invades $j$ because $b_i>b_j$, while $j$ invades $i$ provided that
\begin{equation}
b_jc_jR_j-b_i c_i A_i>0. \label{eq:jinvadcoex}
\end{equation}
Thus, coexistence occurs if $c_j/c_i$ is large enough. This is a version of the classic competition-colonization trade-off \citep{tilman_94,levins_71}: the competitor ($c$-specialist) leaves many territories unoccupied (low $L$) due to its poor colonization ability (low $b$), which the colonizer ($b$-specialist) can then exploit. A similar argument applies for coexistence between high-$c$ and low-$d$ specialists; a ``competition-longevity'' trade-off \citep{tilman_94}. Mutual invasibility is not possible between $b$- and $d$-specialists.

@@

The classical population genetic treatment of fluctuating selection suggests that environmental fluctuations do not promote coexistence. Allele frequencies are successively multiplied by relative fitness values for each environmental iteration, and so two alleles favored in different enviroments can only stably coexist if the product of fitnesses for one type exactly equals the product for the other \citep{dempster_1955}. Thus, stable coexistence still requires frequency dependent selection or heterozygote advantage (as is required in a constant environment).

This classical argument overlooks two general mechanisms by which fluctuating selection promotes coexistence. The first is the ``storage effect'', which introduces a form of frequency dependent selection that promotes coexistence in the presence of environmental fluctuations but not in a constant environment. The storage effect occurs when some individuals are protected from selection; in the lottery model a fraction $(1-d_i)n_i$ of each type's adults do not experience selection in a given iteration. Protection from selection promotes coexistence in fluctuating environments because abundant types cannot fully exploit environmental periods that favor them (since only a fraction of the rare type can be displaced), whereas rare types gain the full benefits of their favorable periods (far more adults from the abundant type die than can possibly be replaced by the rare types) \citep{chesson_1981}.

The second mechanism we will call the ``bounded density effect'', since it is a consequence of the inhibition of reproduction at high population densities (Dempster's (1955) argument ignores density-dependent effects). If there is growth from low to high density in each environmental cycle, then types that are abundant determine the time available for growth every cycle, with less time spent growing in cycles where they are favored, and more in cycles where they are not. This promotes coexistence even in the absence of frequency dependent selection \citep{yi_2013}.

@@

Neither type has an advantage over a full environmental cycle, and they stably coexist. This is due to some combination of the storage and bounded population size effects (stable coexistence between $b$ and $d$ specialists was not possible in a constant environment). It is clear that the classic lottery model, which has a storage effect but no bounded population size effect, will give very different coexistence predictions from our extension of it, because population size will immediately return to capacity $N=T$ in the first summer iteration, after which type frequencies remain constant until the winter. The $d$ specialist thus effectively has infinitely many propagules to secure its winter frequency gains, an enormous advantage compared to the finite propagule density dynamics in Fig. \ref{fig:fluctuatingselection}. Similar difficulties arise in previous models of how the storage effect promotes genetic variation \citep{ellner_1994}, which assume that the total number of offspring per iteration is constant. Beyond this observation, disentangling the storage and bounded population size effects is not straightforward, and requires a more detailed discussion of each effect than we have space for here. Our model is well suited for such a disentangling since it extends the canonical example of the storage effect to allow for density dependent effects.

>>>>

We assume that the $n_i$ and $T$ are large enough that stochastic fluctuations in the $n_i$ (``drift'') can be ignored. We derive deterministic equations for the expected change in the $n_i$ over time, leaving the evaluation of drift for future work. In particular, we do not evaluate the initial stochastic behaviour of adaptive mutant lineages while they are at low abundance. When considering mutations, we restrict our attention to the earliest deterministic behavior of mutant lineages while they are still at low frequency (the transition to deterministic growth occurs at an abundance of order equal to their inverse expected absolute growth rate; \citealt{uecker_2011}).

@@

Stressful environments are more ambiguous, and have been the subject of an extensive debate in the plant ecology literature (the ``Grime-Tilman'' debate; \citealt{aerts_1999} and references therein). Severe stress inhibits growth and reproduction, so that $b\ll 1$ \citep{grime_1974,grime_1977}. In Grime's view, this means that the rate at which propagules successfully develop to adulthood cannot appreciably exceed the mortality rate $b/d\approx 1$. In our model, this implies low propagule density and little territorial contest.

The alternative view is that stressed environments are highly competitive. In spite of inhibited growth and reproduction, population density $N/T$ is actually high since stressful environments also have lower environmental carrying capacity \citep{taylor_1990}. For example, if consumable resources are scarce, we expect intense resource competition (for empirical support, see \citealt{davis_1998}). In our model, territorial contests will then be important.

The mapping of different environments to our model parameters is summarized in the first two rows of Fig.~\ref{fig:table}. Also shown is the approximate dependence of $\Delta_+ n_i$ on $b_i$ and $c_i$ for each environment (fourth row; details in Appendix). These can be used infer the expected direction of evolution for the traits $b$, $c$ and $d$ (fifth row) using a standard invasion analysis.

According to Haldane's formula, the probability that an adaptive mutant survives its initial low abundance phase and begins to grow deterministically is proportional to its expected absolute growth rate, with a proportionality factor typically of order one (\citealt{uecker_2011}) [find Haldane reference]. The fixation of neutral mutations is exceedingly unlikely (probability of order $1/N$). Consequently, the direction of evolutionary change is determined by the mutational trait changes which are available and also confer an appreciable fitness benefit, where availability is subject to constraints imposed by the environment.

@@

Stable coexistence is possible between genotypes that are superior in different traits. Suppose that $j$ is better at securing territories ($c_j>c_i$), that $i$ is better at producing propagules ($b_i>b_j$), and that $d_i=d_j$. Coexistence occurs if $j$ will invade an $i$-dominated population, but $i$ will also invade a $j$-dominated population (``mutual invasion''). If $b_i$ is so large that $L\gg 1$ when $i$ is dominant, and $b_j$ is so small that $L\ll 1$ when $j$ is dominant, then, combining Eqs. \eqref{eq:equil} and \eqref{eq:invad}, we find that $i$ invades $j$ because $b_i>b_j$, while $j$ invades $i$ provided that
\begin{equation}
b_jc_jR_j-b_i c_i A_i>0. \label{eq:jinvadcoex}
\end{equation}
Thus, coexistence occurs if $c_j/c_i$ is large enough. This is a version of the classic competition-colonization trade-off \citep{tilman_94,levins_71}: the competitor ($c$-specialist) leaves many territories unoccupied (low $L$) due to its poor colonization ability (low $b$), which the colonizer ($b$-specialist) can then exploit. A similar argument applies for coexistence between high-$c$ and low-$d$ specialists; a ``competition-longevity'' trade-off \citep{tilman_94}. These forms of co-existence require density dependence, and are not present in the classic lottery model. Mutual invasibility is not possible between $b$- and $d$-specialists.

@@

The classical population genetic treatment of fluctuating selection suggests that environmental fluctuations do not promote coexistence. Allele frequencies are successively multiplied by relative fitness values for each environmental iteration, and so two alleles favored in different enviroments can only stably coexist if the product of fitnesses for one type exactly equals the product for the other \citep{dempster_1955}. Thus, stable coexistence still requires frequency-dependent selection or heterozygote advantage (as is required in a constant environment).

This classical argument overlooks two effects that are present in our extended lottery model. The first is the storage effect, which requires part of the population to be protected from selection each iteration (as occurs when generations are non-overlapping) \citep{chesson_1981}, and which is also present in the classic lottery model. The second is the  bounded population size effect of \cite{yi_2013}, which requires fluctuations in population density and density-dependent regulation of population growth.

%two general mechanisms by which fluctuating selection promotes coexistence. The first is the ``storage effect'', which introduces a form of frequency dependent selection that promotes coexistence in the presence of environmental fluctuations but not in a constant environment. The storage effect occurs when some individuals are protected from selection; in the lottery model a fraction $(1-d_i)n_i$ of each type's adults do not experience selection in a given iteration. Protection from selection promotes coexistence in fluctuating environments because abundant types cannot fully exploit environmental periods that favor them (since only a fraction of the rare type can be displaced), whereas rare types gain the full benefits of their favorable periods (far more adults from the abundant type die than can possibly be replaced by the rare types) \citep{chesson_1981}.

%The second mechanism we will call the ``bounded density effect'', since it is a consequence of the inhibition of reproduction at high population densities (Dempster's (1955) argument ignores density-dependent effects). If there is growth from low to high density in each environmental cycle, then types that are abundant determine the time available for growth every cycle, with less time spent growing in cycles where they are favored, and more in cycles where they are not. This promotes coexistence even in the absence of frequency-dependent selection \citep{yi_2013}.

@@

Neither type has an advantage over a full environmental cycle, and they stably coexist. This is due to some combination of the storage and bounded population size effects (stable coexistence between $b$ and $d$ specialists was not possible in a constant environment). The classic lottery model gives dramatically different co-existence predictions in this scenario (keeping all other parameters the same, the birth rate of the winter survivor must be about $5$ times smaller). This occurs because density returns immediately to capacity $N=T$ in the first summer iteration, after which type frequencies remain constant until the winter. The $d$ specialist thus effectively has infinitely many propagules to secure its winter frequency gains, an enormous advantage compared to the finite propagule density dynamics in Fig. \ref{fig:fluctuatingselection}. Similar difficulties arise in previous models of how the storage effect promotes genetic variation \citep{ellner_1994}, which assume that the total number of offspring per iteration is constant. Beyond this observation, disentangling the storage and bounded population size effects is not straightforward, and requires a more detailed discussion of each effect than we have space for here. Our model is well suited for such a disentangling since it extends the canonical example of the storage effect to allow for density dependent effects.
